Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-13 RNARedPrint-Bioinfo2018.tex
Shorten acknowledgments etc (minor)

To illustrate our above strategy, we sampled $n=1\,000$ sequences
targeting ${\sf GC}_{target}=0.5$ and different Turner energies for
the three structure targets of an example
instance. Figure~\ref{fig:energydist}A illustrates how the Turner
energy distributions of the shown target structures can be accurately
shifted to prescribed target energies; for comparison, we plot the
respective energy distribution of uniformly and Boltzmann sampled
sequences. Figure~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
are directly targeted (at $\pm$10\% tolerance, expect for a few hard
instances). Due to our linear adjustment, we still achieve well
controlled energies in the ultimately relevant Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
  \end{center}
  \caption{%
  	Targeting specific energies using multi-dimensional Boltzmann sampling. (A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors. For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$. (B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath}, \texttt{3str} and \texttt{4str}. Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
}
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.

@@

Using our strategy, we generated at least 1\,000 seeds per instance of
the subsets \texttt{RNAtabupath}, \texttt{3str}, and \texttt{4str}, of
the \Software{Modena} benchmark set with 2,3, and 4 structures. The
results are compared, in terms of their value for the objective
function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences
uniformly sampled using \Software{RNAblueprint}~\cite{Hammer2017}
(Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
that Boltzmann-weighted sequences constitute better seeds, associated
with better values ($\sim\! 69\%$ improvement on average), compared to
uniform seeds for \emph{all instances} of our benchmark
(Supp. section~\ref{sec:validity}).

@@

\begin{table}[t]
\centering
\medskip
\begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
             &   Dataset   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
  Seeds      & RNAtabupath & 21.67 ($\pm$4.38) & 37.74 ($\pm$6.45) & 73\%\\
             & 3str & 18.09 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\\midrule

  Optimized  & RNAtabupath & 5.84 ($\pm$1.31) & 7.95 ($\pm$1.76) & 28\%\\
             & 3str & 5.08 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
             & 4str & 8.77($\pm$1.48) & 13.13 ($\pm$2.13) & 37\% \\ \bottomrule
\end{tabular}\\[1em]

@@

\newpage 
\section*{Acknowledgements}
YP is supported by the French/Austrian RNALands project (ANR-14-CE34-0011), jointly supported by the French {\em Agence Nationale de la Recherche} and the Austrian FWF.
SH is supported within the
German Network for Bioinformatics Infrastructure (de.NBI) by the
German Federal Ministry of Education and Research (BMBF; support code
031A538B) and the project RiboNets acknowledged by financial support of the Future and Emerging Technologies (FET) programme within the Seventh Framework Programme for Research of the European Commission, under FET-Open grant number: 323987. 
All authors wish to express their gratitude to Leonid Chindelevitch for very accurately pointing out that the bipartite nature of the dependency graph directly implies a reduced complexity for our dynamic programming scheme. We also thank Arie Koster for fruitful discussions regarding the practical computation of the tree decomposition.

>>>>

To illustrate our above strategy, we sampled $n=1\,000$ sequences
targeting ${\sf GC}_{target}=0.5$ and different Turner energies for
the three structure targets of an example
instance. Figure~\ref{fig:energydist}A illustrates how the Turner
energy distributions of the shown target structures can be accurately
shifted to prescribed target energies; for comparison, we plot the
respective energy distribution of uniformly and Boltzmann sampled
sequences. Figure~\ref{fig:energydist}B summarizes the targeting accuracy for the single structure energies over a larger set of instances
from the \texttt{Modena} benchmark. We emphasize that only the simple stacking energies
are directly targeted (at $\pm$10\% tolerance, expect for a few hard
instances). Due to our linear adjustment, we still achieve well
controlled energies in the ultimately relevant Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
    {\sf \bfseries B}\includegraphicstop[width=0.32\textwidth]{Figs/offset}
  \end{center}
  \caption{%
  	Targeting specific energies using multi-dimensional Boltzmann sampling. (A) Turner energy distributions for three target structures (right) while targeting $(-40,-40,-20)$ and $(-20,-20,-20)$ free energy vectors. For comparison, we show the distribution of uniform and Boltzmann samples; respectively associated with homogenous weights $1$ and $e^\beta$. (B) Accuracy of targeting over all target structures of the \Software{Modena} benchmark instances \texttt{RNAtabupath} (2str), \texttt{3str} and \texttt{4str}. Shown is the relative deviation of targeted and achieved energies in the simple stacking model and the Turner model.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
}
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenario.

@@

Using our strategy, we generated at least 1\,000 seeds per instance of
the subsets \texttt{RNAtabupath} (2str), \texttt{3str}, and \texttt{4str}, of
the \Software{Modena} benchmark set with 2,3, and 4 structures. The
results are compared, in terms of their value for the objective
function of Eq.~\eqref{eq:blueprintobjective}, to seed sequences
uniformly sampled using \Software{RNAblueprint}~\cite{Hammer2017}
(Table~\ref{tab:benchmark-results}, seeds). Our first analysis reveals
that Boltzmann-weighted sequences constitute better seeds, associated
with better values ($\sim\! 69\%$ improvement on average), compared to
uniform seeds for \emph{all instances} of our benchmark
(Supp. section~\ref{sec:validity}).

@@

\begin{table}[t]
\centering
\medskip
\begin{tabular}{@{}>{\bf}l@{\quad}>{\tt}l@{\quad}@{\quad}c@{\quad}c@{\quad}c@{}}
             &   \textbf{\textrm{Dataset}}   & {\bfseries\ourprog{}} & \textbf{Uniform} & \textbf{Improvement} \\\toprule
  Seeds      & 2str & 21.67 ($\pm$4.38) & 37.74 ($\pm$6.45) & 73\%\\
             & 3str & 18.09 ($\pm$3.98) & 30.49 ($\pm$5.41) & 71\%\\
             & 4str & 19.94 ($\pm$3.84) & 32.29 ($\pm$5.24) & 63\%\\\midrule

  Optimized  & 2str & 5.84 ($\pm$1.31) & 7.95 ($\pm$1.76) & 28\%\\
             & 3str & 5.08 ($\pm$1.10) & 7.04 ($\pm$1.52) & 31\%\\
             & 4str & 8.77($\pm$1.48) & 13.13 ($\pm$2.13) & 37\% \\ \bottomrule
\end{tabular}\\[1em]

@@

\newpage 
\section*{Acknowledgements}
YP is supported 
%jointly supported 
by
% the French 
{\em Agence
  Nationale de la Recherche} and the 
%Austrian
FWF (ANR-14-CE34-0011; project RNALands).  SH is supported by the
German Federal Ministry of Education and Research (BMBF support code
031A538B; de.NBI: German Network for Bioinformatics Infrastructure)
and the
% acknowledged by
%financial support 
of the Future and Emerging Technologies programme
% within the Seventh Framework Programme for Research of the
%European Commission, under 
(FET-Open grant 323987; project RiboNets).
%
We thank Leonid Chindelevitch for pointing out the potential for
complexity reductions due to the bipartiteness of the compatibility
graph; moreover, Arie Koster for discussing the practical computation
of the tree decomposition.
